---
title: "Assignment #1"
author: "Jiaqi Ma"
date: "2026-02-06"
output:
  html_document:
    toc: true                   # this is to generate a table of contents
    toc_depth: 2                # this is to control what level of header we are going to display on the table of contents 
    toc_float: true             #this is to have a floating table of contents
    code_folding: none          # this is to fold/hide the code 
bibliography: BCB420.bib        # this is for the citations/bibliography 
link-citations: true            # this is for the citations/bibliography
---

# Data Download

------------------------------------------------------------------------

-   My choice of dataset is `GSE209552`.

```{r}
# Define global constant
GSE_NUM <- "GSE209552"
```

## GEOmetadb setup

To setup [GEOmetadb Tutorial](https://www.bioconductor.org/packages/release/bioc/html/GEOmetadb.html), follow [GEOmetadb Tutorial Chapter 3](https://bcb420-2026.github.io/Exercise_Finding_Expression_data/setting-up-geometadb.html)

```{r}
# Install required package if not installed
if (!requireNamespace("BiocManager", quietly = TRUE))
  install.packages("BiocManager")
if (!requireNamespace("GEOmetadb", quietly = TRUE))
  BiocManager::install("GEOmetadb")
if (!requireNamespace("DBI", quietly = TRUE))
  install.packages("DBI")
if (!requireNamespace("RSQLite", quietly = TRUE))
  install.packages("RSQLite")
if (!requireNamespace("dplyr", quietly = TRUE))
  install.packages("dplyr")
if (!requireNamespace("ggplot2", quietly = TRUE))
  install.packages("ggplot2")

# Attach packages
library(GEOmetadb)
library(DBI)
library(RSQLite)

# Download SQLite database if not exist
if (!file.exists('GEOmetadb.sqlite')) getSQLiteFile()
file.info('GEOmetadb.sqlite')

# Connect to database
con <- dbConnect(SQLite(), 'GEOmetadb.sqlite')
```

## Download whole GSE209552

Download data, follow [Identifier Mapping Tutorial](https://bcb420-2026.github.io/Example_Student/id_mapping/getting-data-from-geo.html). The helper function defined below is modified based on [Identifier Mapping Tutorial's helper function](https://github.com/bcb420-2026/Example_Student/blob/main/in_class_exercises/id_mapping/fetch_geo_supp.R) to avoid duplicated download.

```{r}
# Define helper function
fetch_geo_supp <- function(gse, destdir = "data") {
  # Calculate the actual path where GEOquery puts the files
  target_path <- file.path(destdir, gse)
  
  # Check if directory exists AND contains files
  if (dir.exists(target_path) && length(list.files(target_path)) > 0) {
    message(paste("Data found at", target_path, "- Skipping download."))
    return(invisible(target_path))
  }
  
  # If missing or empty, proceed with download
  message(paste("Downloading", gse, "to", destdir, "..."))
  dir.create(destdir, showWarnings = FALSE, recursive = TRUE)
  GEOquery::getGEOSuppFiles(GEO = gse, baseDir = destdir, makeDirectory = TRUE)
  
  invisible(target_path)
}

# Download
fetch_geo_supp(gse = GSE_NUM)
```

## Construct scRNA-seq only reference list

This GSE contain both scRNA-seq and bulk RNA-seq sample. Removal of bulk RNA-seq is needed to avoid significant batch effect due to their technical difference. Although the number of sample is reduced to 17, it's still sufficient amount.

Refer to the [original paper](https://doi.org/10.1016/j.celrep.2023.113395)'s method section, only the scRNA-seq was done by `Illumina NextSeq6000` platform, [GPL24676](https://www-ncbi-nlm-nih-gov.myaccess.library.utoronto.ca/geo/query/acc.cgi?acc=GPL24676). We can use this clue to exclude all bulk RNA-seq

Since all GSMs of this study are packed in one `GSE209552_RAW.tar` file in GEO, we can't indicate which platform to include/exclude at download step. Thus we will construct a list of `GPL24676` GSMs for downstream reference by query `GEOmetabd`.

```{r}
# Define desired platform
scRNA_platform <- "GPL24676"

# Construct the query
# We join the 'link' table (gse_gsm) with the 'sample' table (gsm) 
# to check the platform (gpl) column.
sql_filter <- paste0(
  "SELECT t1.gsm ",
  "FROM gse_gsm t1 ",
  "JOIN gsm t2 ON t1.gsm = t2.gsm ",
  "WHERE t1.gse = '", GSE_NUM, "' ",
  "AND t2.gpl = '", scRNA_platform, "'"
)

# Get the list of pure scRNA-seq samples
sc_gsm_list <- dbGetQuery(con, sql_filter)$gsm

# View results
print(paste("Found", length(sc_gsm_list), "scRNA-seq samples."))
sc_gsm_list
```

# Access data quality

-   The following analysis is build on [Seurat](https://satijalab.org/seurat/) framework and referring [Heumos et al, 2023](https://doi.org/10.1038/s41576-023-00586-w) for high-level idea.

```{r}
# Install Seurat if not
if (!requireNamespace("Seurat", quietly = TRUE))
  install.packages("Seurat")
```

## Reorganize scRNA-seq files

```{r}
# Attach packages
library(Seurat)
library(data.table)
library(Matrix)
library(dplyr)
library(stringr)

# Define data paths
tar_file <- "data/GSE209552/GSE209552_RAW.tar"
extract_dir <- "data/GSE209552/raw_files"

# Define unpack helper function
unpack_geo_tar <- function(tar_path, dest_dir) {
  if (dir.exists(dest_dir) && length(list.files(dest_dir)) > 0) {
    message(paste("📂 Extracted files found at", dest_dir, "- Skipping unpack."))
    return(invisible(dest_dir))
  }
  if (!file.exists(tar_path)) {
    stop(paste("❌ Tar file not found:", tar_path))
  }
  message(paste("📦 Unpacking", basename(tar_path), "..."))
  dir.create(dest_dir, showWarnings = FALSE, recursive = TRUE)
  untar(tar_path, exdir = dest_dir)
  message("✅ Unpacking complete.")
  return(invisible(dest_dir))
}

# Run the unpacker
unpack_geo_tar(tar_file, extract_dir)

# Reorganize all scRNA-seq 10x files in to a new directory
dest_dir <- "data/GSE209552/10x_organized"    # Where they should go
dir.create(dest_dir, recursive = TRUE, showWarnings = FALSE)

# --- MAIN LOOP ---
message(paste("🚀 Organizing files for", length(sc_gsm_list), "samples..."))

skipped_count <- 0
moved_count   <- 0

for (gsm in sc_gsm_list) {
  
  # 1. Define the specific destination path for this sample
  sample_subdir <- file.path(dest_dir, gsm)
  
  # 2. CHECK: Does it already exist?
  if (dir.exists(sample_subdir) && length(list.files(sample_subdir)) >= 3) {
    skipped_count <- skipped_count + 1
    next
  }
  
  # 3. Create directory
  dir.create(sample_subdir, showWarnings = FALSE)
  
  # 4. Find source files (Exclude the 'TE' files)
  pattern <- paste0("^", gsm)
  files_to_move <- list.files(extract_dir, pattern = pattern, full.names = TRUE)
  files_to_move <- files_to_move[!grepl("_TE_", files_to_move)]
  
  # Safety Checks
  if (length(files_to_move) == 0) {
    warning(paste("⚠️ Skipping", gsm, "- No source files found."))
    next
  }
  
  # 5. MOVE & RENAME (The Standardize Step)
  # We identify specific files to ensure we rename them correctly for Seurat
  f_matrix   <- files_to_move[grep("matrix", files_to_move, ignore.case = TRUE)]
  f_barcodes <- files_to_move[grep("barcodes", files_to_move, ignore.case = TRUE)]
  f_features <- files_to_move[grep("features|genes", files_to_move, ignore.case = TRUE)]
  
  # Verify we found exactly 1 of each before moving
  if (length(f_matrix) == 1 && length(f_barcodes) == 1 && length(f_features) == 1) {
    
    # Rename Matrix -> matrix.mtx.gz
    file.rename(f_matrix, file.path(sample_subdir, "matrix.mtx.gz"))
    
    # Rename Barcodes -> barcodes.tsv.gz
    file.rename(f_barcodes, file.path(sample_subdir, "barcodes.tsv.gz"))
    
    # Rename Genes/Features -> features.tsv.gz
    # (Note: Seurat prefers 'features.tsv' over 'genes.tsv')
    file.rename(f_features, file.path(sample_subdir, "features.tsv.gz"))
    
    message(paste("  ✅ Standardized & Moved:", gsm))
    moved_count <- moved_count + 1
    
  } else {
    warning(paste("  ⚠️ Skipping", gsm, "- Could not find unique 10x triplet to standardize."))
  }
}

# --- SUMMARY ---
message("🎉 Organization Complete.")
message(paste("  🔹 Samples moved & standardized:", moved_count))
message(paste("  🔸 Samples skipped (already done):", skipped_count))
```

## Load scRNA-seq files into R using Seurat

```{r}
# Get the list of sample subdirectories
sample_dirs <- list.dirs(dest_dir, recursive = FALSE)

# Define a list storing seurat obj of all GSMs
seurat_list <- list()

message(paste("🚀 Loading", length(sample_dirs), "samples..."))
for (dir in sample_dirs) {
  # Extract GSM ID from the folder name (e.g., "GSM6385438")
  gsm_id <- basename(dir)
  
  # A. Read 10x Data
  # Because you renamed the files, Read10X finds them automatically!
  counts <- Read10X(data.dir = dir)
  
  # B. Create Seurat Object
  seurat_list[[gsm_id]] <- CreateSeuratObject(
    counts = counts, 
    project = gsm_id, 
    min.cells = 0, 
    min.features = 0
  )
  
  message(paste("  ✅ Loaded:", gsm_id))
}
length(seurat_list)
```

## Data exploriation

Plot the following statistics:

-   Number of unique identifier among cells (infer coverage)
-   Number of total identifier among cells (infer depth)
-   Number of samples and cells in control group
-   Number of samples and cells in treatment group

```{r}
library(ggplot2)
library(patchwork)

gsm_list_string <- paste(paste0("'", sc_gsm_list, "'"), collapse = ",")
sql <- paste0(
  "SELECT gsm, title, characteristics_ch1 ",
  "FROM gsm ",
  "WHERE gsm IN (", gsm_list_string, ")"
)
metadata <- dbGetQuery(con, sql)
```

